---
title: "M01-2-Introduction to Literate Programming-Application Assignment"
author: "Claudia Flores"
date: '`r format(Sys.time())`'
format: 
  html: 
    toc: true
    toc-depth: 4
    number-sections: true
    code-line-numbers: true
    code-fold: false
    code-link: true
    embed-resources: true
editor: visual
execute: 
  freeze: auto
  warning: false
---

# Exercise 

```{r}
X <- "This is my first assignment"
```

X \<- "This is my first assignment"

# Exercise 

## Exercise:

```{r}
paste("Good", "morning")
paste("Good", "morning", sep = "-")
paste0("Good", "morning")
paste("The value of X is:", X)

```

## Exercise:

```{r}
X <- paste(X, "and I'm loving it")
```

# Exercise 

```{r}
Y <- c(2,3,4,5)
Y <- Y*2
```

# Exercise 

```{r}
X
Y
```

# Exercise 

```{r}
max(Y)
min(Y)
```

# Exercise 

```{r}
# install.packages("ggplot2")
library(ggplot2)
head(economics)
```

# Exercise 

```{r}
plot <- ggplot(
  data = economics,
  mapping = aes(x = unemploy, y = uempmed)
) +
  geom_point()

plot +
  labs(
    title = "Unemployment and Median Duration of Unemployment",
    x = "Number of Unemployed People in Thousands",
    y = "Median Duration of Unemployment in Weeks"
  )
```
# Exercise 

```{r}
#| label: fig-unemployment
#| fig-cap: "Relationship between number of unemployed people and median duration of unemployment."
#| fig-width: 5
#| fig-asp: 0.6

economics |>
  ggplot(aes(x = unemploy, y = uempmed)) +
  geom_point() +
  labs(
    title = "Unemployment and Median Duration of Unemployment",
    x = "Number of Unemployed People in Thousands",
    y = "Median Duration of Unemployment in Weeks"
  )
```
# Exercise
:: {.callout-important}

In @fig-unemployment, we see a positive connection between number of unemployed people and the median duration of unemployment. As the number of unemployed people increases, the median duration of unemployment increases as well.

:::

# Exercise

## List of tools I used   
  1.  Section numbering with "\#", "\##", and "\###".  
  2.  bold font for names of variables. 
  3.  italicized font for certain variables.  
  4.  `callout-tip` for all hints.  
  5.  `callout-note` for information. 
  6.  `callout-important` for what we found out. 
  7.  `blockquote` for tasks. 
  8.  tick ("\`") before and after a function to set it apart from normal texts.
  9.  cross-reference the line figure by using @fig-unemployment. 
  10. Optional chunk argument (`#|`) inside some code chunks to control code chunks. 
  